---
title: "Classroom Notes, Week 2 - Exploratory Data Analysis"
author: "Reginald Carey"
date: "March 6, 2015"
output:
  html_document: default
---

Lattice Plotting System

xyplot

bwplot

histogram

stripplot

dotplot

splom

levelplot, contourplot

Lattice functions generally take a function as the first parameter

```
xyplot(y ~ x | f * g, data)
```

on the left of the ~ is the y-axis, on the right is the x-axis

f and g are optional conditioning variables, the * indicates an interaction bewteen two variables.

data is a frame or list.  If no data frame is passed, then the parent frame is used.

# Simple Lattice Plot

```{r,cache=TRUE}
library(datasets)
library(lattice)
## Convert 'month' to a factor variable
airquality <- transform(airquality, Month = factor(Month))
xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5, 1))
```

Lattice produces objects of type *trellis*

autoprinting takes care of printing a *trellis* object (creates the graph)

# Lattice Behaviour

```{r,cache=TRUE}
p <- xyplot(Ozone ~ Wind, data = airquality) ## Nothing happens here!
```

Now we try to print it:

```{r,cache=TRUE}
print(p) ## Plot appears
```

Auto printing:
```{r,cache=TRUE}
xyplot(Ozone ~ Wind, data = airquality) ## Auto-printing
```

--------

# Lattice Panel Functions

```{r,cache=TRUE}
set.seed(10)
x <- rnorm(100)
f <- rep(0:1, each = 50)
y <- x + f - f * x  + rnorm(100, sd = 0.5)
f <- factor(f, labels = c("Group 1", "Group 2"))
xyplot(y ~ x | f, layout = c(2,1))  ## Plot with 2 panels
```


Custom Panel Function

```{r,cache=TRUE}
xyplot(y ~ x | f, panel = function(x,y,...) {
  panel.xyplot(x,y,...) ## first call default panel function for 'xyplot'
  panel.abline(h = median(y), lty = 2) ## add a horizontal dashed line at the median
})
```

Regression Line

```{r,cache=TRUE}
xyplot(y ~ x | f, panel = function(x, y, ...) {
  panel.xyplot(x, y, ...) ## first call default panel function for 'xyplot'
  panel.lmline(x, y, col = 2) ## add a horizontal dashed line at the median
})
```

--------

# GGPLOT2 - part 1

Grammar of Graphics

Documentation at http://ggplot2.org

##  The Basics: qplot()

* Works similar to plot function in base graphics system.

* Looks for data in a data frame, similar to lattice, or in the parent environment.

* plots are made up of aesthetics (size, shape, color) and geoms (points, lines)

* Factors are important because they indicate subsets of the data.

* The qplot() hides what goes on underneath.

* ggplot() is the core function and very flexible for doing things qplot() cannot do.

--------

# GGPLOT - part 2

```{r,cache=TRUE}
library(ggplot2)
str(mpg)
```

### ggplot2 "Hello, world!"

```{r,cache=TRUE}
library(ggplot2)
qplot(displ, hwy, data = mpg)
```

### Modifying the aesthetics

```{r,cache=TRUE}
library(ggplot2)
qplot(displ, hwy, data = mpg, color = drv)
```

### Adding a geom - loess line

```{r,cache=TRUE}
library(ggplot2)
qplot(displ, hwy, data = mpg, geom=c("point", "smooth"), method="loess")
```

### Histograms

```{r,cache=TRUE}
library(ggplot2)
qplot(hwy, data = mpg, fill = drv)
```

### Facets

```{r,cache=TRUE}
library(ggplot2)
qplot(displ, hwy, data = mpg, facets = . ~ drv)
```

```{r,cache=TRUE}
library(ggplot2)
qplot(hwy, data = mpg, facets = drv ~ ., binwidth = 2)
```

--------

# The ggplot2 Plotting System - Part 2

An implementation of the *Grammar of Graphics* by Leland Wilkinson

## Basic components

* A **data frame**
* **aesthetic mappings**: how data are mapped to color, size, etc.
* **geoms**: geometric objects like points, likes, shapes.
* **facets**: for conditional plots.
* **stats**: statistical transformations like binning, quantiles, smoothing.
* **scales**: what scale an aesthetic map users (example: male = red, female = blue).
* **coordinate system**

--------

## Building Plots with ggplot2

* When building plots in ggplot2, the "artist's palette" model may be the closest analogy
* Plots are built up in layers
    + plot the data
    + overlay a summary
    + metadata and annotation

--------

# Quiz # 2

## Question 7

```{r,cache=TRUE}
library(datasets)
data(airquality)
```

```{r,cache=TRUE}
airquality = transform(airquality, Month = factor(Month))
qplot(Wind, Ozone, data = airquality, facets = . ~ Month)
```
