---
title: "Classroom Notes Week 3"
author: "Reginald Carey"
date: "March 14, 2015"
output: html_document
---

##Hierarchical Clustering - example

```{r,cache=TRUE}
set.seed(1234)
par(mar = c(0,0,0,0))
x <- rnorm(12, mean = rep(1:3, each = 4), sd = 0.2)
y <- rnorm(12, mean = rep(c(1,2,1), each = 4), sd = 0.2)
plot(x,y,col="blue",pch=19,cex=2)
text(x+0.05, y+0.05, labels = as.character(1:12))
```

```{r,cache=TRUE}
dataFrame = data.frame(x=x,y=y)
dist(dataFrame)
```

```{r,cache=TRUE}
distxy <- dist(dataFrame)
hClustering = hclust(distxy)
plot(hClustering)
```

Heatmap

```{r,cache=TRUE}
dataFrame = data.frame(x=x,y=y)
set.seed(143)
dataMatrix <- as.matrix(dataFrame)[sample(1:12),]
heatmap(dataMatrix)
```

K-Means Clustering

```{r,cache=TRUE}
set.seed(1234)
par(mar = c(0,0,0,0))
x <- rnorm(12, mean = rep(1:3, each = 4), sd = 0.2)
y <- rnorm(12, mean = rep(c(1,2,1), each = 4), sd = 0.2)
plot(x,y,col="blue",pch=19,cex=2)
text(x+0.05, y+0.05, labels = as.character(1:12))
```

```{r,cache=TRUE}
dataFrame = data.frame(x=x,y=y)
kmeansObj <- kmeans(dataFrame, centers = 4)
names(kmeansObj)
```

```{r,cache=TRUE}
kmeansObj$cluster
```


```{r,cache=TRUE}
par(mar = rep(0.2, 4))
plot(x,y, col = kmeansObj$cluster, pch = 19, cex = 2)
points(kmeansObj$centers, col = 1:4, pch = 3, cex = 3, lwd = 3)
```

#Dimension Reduction

```{r,cache=TRUE}
set.seed(12345)
#par(mar = rep(0.2, 4))
dataMatrix <- matrix(rnorm(400), nrow = 40)
image(1:10, 1:40, t(dataMatrix)[, nrow(dataMatrix):1])
heatmap(dataMatrix)
set.seed(678910)
for (i in 1:40) {
    #flip a coin
    coinFlip = rbinom(1, size = 1, prob = 0.5)
    #if coin is heads add a common pattern to that row
    if (coinFlip) {
        dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,3), each = 5)
    }
}
image(1:10, 1:40, t(dataMatrix)[, nrow(dataMatrix):1])
heatmap(dataMatrix)

hh <- hclust(dist(dataMatrix))
dataMatrixOrdered <- dataMatrix[hh$order,]
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[, nrow(dataMatrixOrdered):1])
plot(rowMeans(dataMatrixOrdered), 40:1, xlab = "Row Mean", ylab = "Row", pch = 19)
plot(colMeans(dataMatrixOrdered), xlab = "Column", ylab = "Column Mean", pch = 19)
```

#Components of the SVD - *u* and *v*

```{r,cache=TRUE}
svd1 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[, nrow(dataMatrixOrdered):1])
plot(svd1$u[,1], 40:1, , xlab = "Row", ylab = "First left singular vector", pch = 19)
plot(svd1$v[,1], xlab = "Column", ylab = "First right singular vector", pch = 19)
```


#Components of the SVD - Variance explained

```{r,cache=TRUE}
par(mfrow = c(1,2))
plot(svd1$d, xlab = "Column", ylab = "Singular Value", pch = 19)
plot(svd1$d^2/sum(svd1$d^2), xlab = "Column", ylab = "Prop. of variance explained", pch = 19)
```

#Relationship to principal components
```{r,cache=TRUE}
svd1 <- svd(scale(dataMatrixOrdered))
pca1 <- prcomp(dataMatrixOrdered, scale = TRUE)
plot(pca1$rotation[, 1], svd1$v[, 1], pch = 19, xlab = "Principal Component #1", ylab = "Right Singular Vector #1")
abline(c(0,1))
```

#Components of SVD - variance explained
In this plot of the variance we see that 100% of the variance is explained in the first singular value.  That's because 
this matrix is essentially binary in nature.

```{r,cache=TRUE}
constantMatrix <- dataMatrixOrdered * 0
for (i in 1:dim(dataMatrixOrdered)[1]) { constantMatrix[i,] <- rep(c(0,1),each = 5)}
svd1 <- svd(constantMatrix)
par(mfrow = c(1,3))
image(t(constantMatrix)[, nrow(constantMatrix):1])
plot(svd1$d, xlab =  "Column", ylab = "Singular Value", pch = 10)
plot(svd1$d^2/sum(svd1$d^2), xlab = "Column", ylab = "Prop. of variance explained", pch = 19)
```

Lets add a second pattern to the data set

```{r,cache=TRUE}
set.seed(678910)
for (i in 1:40) {
    #flip a coin
    coinFlip1 = rbinom(1, size = 1, prob = 0.5)
    coinFlip2 = rbinom(1, size = 1, prob = 0.5)
    #if coin is heads add a common pattern to that row
    if (coinFlip1) {
        dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5), each = 5)
    }
    if (coinFlip2) {
        dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5), 5)
    }
}
hh <- hclust(dist(dataMatrix))
dataMatrixOrdered <- dataMatrix[hh$order, ]
```

#Singular Value Decomposition - true patterns

```{r,cache=TRUE}
svd2 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[, nrow(dataMatrixOrdered):1])
plot(rep(c(0,1), each = 5), pch = 19, xlab = "Column", ylab = "Pattern 1")
plot(rep(c(0,1), 5), pch = 19, xlab = "Column", ylab = "Pattern 2")
```

#Singular Value Decomposition - *v* and patterns of variance in rows

```{r,cache=TRUE}
svd2 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,3))
image(t(dataMatrixOrdered)[, nrow(dataMatrixOrdered):1])
plot(svd2$v[,1], pch = 19, xlab = "Column", ylab = "First right singular vector")
plot(svd2$v[,2], pch = 19, xlab = "Column", ylab = "Second right singular vector")
```

## *d* and variance explained

```{r}
svd1 <- svd(scale(dataMatrixOrdered))
par(mfrow = c(1,2))
plot(svd1$d, pch = 19, xlab = "Column", ylab = "Singular Value")
plot(svd1$d^2/sum(svd1$d^2), pch = 19, xlab = "Column", ylab = "Percent of variance explained")
```

#Plotting and Color in R

```{r}
smoothScatter(rnorm(10000),rnorm(10000))
```

